### Load Packages

```{r}
#| label: load-pkgs
#| message: false
#| warning: false

library(tidyverse)
library(ggplot2)
```

### Load Data

```{r}
#| label: load-data
#| message: false

wcmatches <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-11-29/wcmatches.csv')
worldcups <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-11-29/worldcups.csv')
penalties <- readr::read_csv("data/penalties_stats_aa.csv") #external data from kaggle
```

#### Merge Data

```{r, echo = FALSE}
#| label: merge-data

world_cup_total <- wcmatches |>
  left_join(worldcups, by = "year")

world_cup_total <- penalties %>%
  select(-Host) %>%
  rename('year' = "Year") %>%
left_join(world_cup_total, penalties, by = 'year')
```

#### New Variables

```{r, echo = FALSE}
#| label: new-variables
world_cup_total <- world_cup_total |>
  mutate(goalspergame = goals_scored/games,
         attendancepergame = attendance/games,
         gamesperteam = games/teams)
```

## Question 1

**scatterplot showing the number of goals scored per game of each world cup**

**x-axis will be `year` and the y-axis will be the average `goals_scored` per game of that world cup**.

**annotations on top of this graph to show when certain changes were implemented in the World Cup**,

**may add another variable such as continent (color) in order to see if the if the power rankings across the world have shifted over time**

```{r, echo = FALSE}
#| label: q1-plot1
world_cup_total %>%
  add_row(year = 1942, goalspergame = 0, country = "Germany") %>%
  add_row(year = 1946, goalspergame = 0, country = "Brazil") %>%
  mutate(new_rule = case_when(year %in% c(1970, 1982, 1998, 2014, 2018) ~ "Yes",
                              TRUE ~ "No"),
         event = case_when(year %in% c(1942, 1946, 1954, 1966, 1978) ~ "Yes",
                           TRUE ~ "No")) %>%
  group_by(year, goalspergame) %>%
  ggplot(aes(x = year, 
             y = goalspergame, 
             color = new_rule, 
             shape = event)) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(1930, 2018, by=8)) +
  theme_minimal() +
  scale_color_manual(values = c("#A80040", "#75A9CF")) +
  scale_shape_manual(values = c(16, 8)) +
  labs(title = "Goals per Game",
       subtitle = "For World Cups 1930-2018",
       x = "Year",
       y = "Goals per game",
       color = "New rule implemented",
       shape = "Notable event") +
  annotate(geom = "text",
           label = "1942, 1946: cancelled \ndue to WWII",
           x = 1944,
           y = 0.5,
           size = 2.5) +
  annotate(geom = "text",
           label = "1954: most goals by\n1 country (Hungary, 27)",
           x = 1949,
           y = 5,
           size = 2.5,
           color = "#A80040") +
  annotate(geom = "text",
           label = "1966: first hattrick\nin a World Cup final",
           x = 1958,
           y = 2.4,
           size = 2.5) +
  annotate(geom = "text",
           label = "1978: hardest World\nCup to qualify for",
           x = 1982,
           y = 5.1,
           size = 2.5, 
           color = "#A80040") +
  geom_segment(aes(x = 1978, y = 2.68,
                   xend = 1978, yend = 4.8),
               color = "#A80040")
```

## Question 2

```{r}
q2 <- wcmatches %>%
  pivot_longer(c(home_team, away_team)) %>%
  rename(home_away = name,
         team = value)

```

### Add Continent

#### Making continent lists

```{r}
Europe  <- c('France', "Belgium", 'Yugoslavia', 'Romania', 'Sweden', 'Austria', 'Germany', 'Spain', 'Czechoslovakia', 'Hungary', 'Italy', 'Netherlands', 'Switzerland', 'Poland', 'Norway', 'England', 'Scotland', 'West Germany', 'Turkey', 'Northern Ireland', 'Wales', 'Bulgaria', 'Portugal', 'East Germany', 'Denmark', 'Republic of Ireland', 'Greece', 'Croatia', 'FR Yugoslavia', 'Serbia', 'Slovenia', 'Czech Republic', 'Ukraine', 'Slovakia', 'Iceland', 'Bosnia and Herzegovina') 

Asia <- c('South Korea', 'Soviet Union', 'North Korea', 'Israel', 'Iran', 'Cameroon', 'Kuwait', 'Iraq', 'United Arab Emirates', 'Russia', 'Saudi Arabia', 'Japan', 'China PR', 'Australia', 'New Zealand')

Africa <- c('Egypt', 'Morocco', 'Zaire', 'Tunisia', 'Algeria', 'Nigeria', 'South Africa', 'Senegal', 'Ivory Coast', 'Angola', 'Ghana', 'Togo' )

North_America <- c('Mexico', 'United States', 'Cuba', 'Dutch West Indies', 'Haiti', 'El Salvador', "Honduras", 'Canada', 'Costa Rica', 'Jamaica', 'Trinidad and Tobago', 'Panama')

South_America <- c('Brazil', 'Peru', 'Argentina', 'Chile', 'Bolivia', 'Paraguay', 'Uruguay', 'Colombia','Ecuador')

```

### Make continent variable

```{r}
q2 <- q2 %>%
  mutate(continent =
           case_when(team %in% Europe ~ 'Europe',
                     team %in% Asia ~ 'Asia/Oceania',
                     team %in% South_America ~ 'South America',
                     team %in% North_America ~ 'North America',
                     team %in% Africa ~ 'Africa',
                     TRUE ~ 'Other'))

```

### Early graphs

```{r}
continent_per_year <- q2 %>%
  group_by(year, continent) %>%
  summarize(teams = unique(team)) %>%
  group_by(year, continent) %>%
  summarize(count_per_continent = n())

freq <- continent_per_year %>%
  group_by(year) %>%
  mutate(total = sum(count_per_continent),
         pct = count_per_continent/total) 
```

```{r}
ggplot(freq, aes(x = year, y = pct, color = continent)) + geom_line()
```
